An Interpretable Machine Learning Framework for Modeling High-Resolution Spectroscopic Data\footnote{Open source code at \url{https://github.com/gully/blase}}

### 2022-09-26 paper/paper1/ms.tex
Address the referee comments on PyTorch ecosystem

A key new enabling technology breaks these classical tradeoffs in data-model comparisons for astronomical spectroscopy.  Automatic differentiation \citep[``autodiff'' or ``autograd'',][]{2015arXiv150205767G, 2016PhDT.......317M} and its affiliated backpropagation algorithm \citep{kelley1960,Linnainmaa1976TaylorEO, 1986Natur.323..533R, 1990JGCD...13..926D} has revolutionized machine learning and neural network architecture design, and is increasingly applied in astrophysical data analysis contexts, \emph{e.g.} kernel phase coronography with \texttt{morphine} \citep{2021ApJ...907...40P,2021JOSAB..38.2465W} \edit1{and $\partial$\texttt{lux} \citep{10.1117/12.2629774}}, and exoplanet orbit fitting with \texttt{exoplanet} \citep{2021JOSS....6.3285F}.  Of the spectroscopy frameworks mentioned above, the \texttt{TensorFlow}-based \citep{tensorflow2015-whitepaper} \texttt{wobble} and the \texttt{JAX}-based \citep{jax2018github} \texttt{ExoJAX} employ autodiff technology.  \texttt{wobble} treats each pixel as a tunable control point, producing $\sim10^5$ parameters for a modern stellar spectrum.  The \texttt{ExoJAX} framework has only $\sim$dozens of tunable parameters that describe the fundamental physical properties controlling a brown dwarf atmosphere.  These two autodiff-aware frameworks span the extreme ends of non-parametric and parametric modeling for spectroscopy.

@@

We initialize the cloned model with a series of preprocessing steps. We divide the entire spectrum by a black body $\mathsf{B}(\bm{\lambda}_S)$ of the same effective temperature $T_{\mathrm{eff}}$ as the model template. The resulting signal usually still has smooth wiggles around the continuum. An optional continuum flattening step ensures that subsequent spectral line finding steps get applied uniformly. This high-pass filtering step should be set to capture the genuine spectral shape, without over-fitting broad line wings such as those in deep hydrogen and sodium lines. Any high-pass filter will work, a Gaussian Process approach would be ideal.  Instead we apply a simple and familiar heuristic: fit a polynomial $\mathsf{P}(\bm{\lambda}_S)$ to a few continuum peaks and divide out the trend.

@@

\emph{``Surrogate modeling / Simulation Intelligence''}-- \edit1{Our slightly off-label use of machine learning straddles a boundary of machine learning and conventional physics-based models.  The rapidly evolving field of machine learning may therefore classify the approach here using a few related terms.  The phrases surrogate modeling, emulation, discrepancy modeling, transfer learning, simulation intelligence, and physics-informed machine learning could describe some aspects of \emph{blas\'e}, depending on the ultimate application by the end-user.  For example, machine learning has been applied to physical chemistry models using a neural network as a cheaper-but-still-accurate ``surrogate'' for the real theory-based simulations. The realized speedups in quantum chemistry can approach 300,000$\times$ \citep{pmlr-v70-gilmer17a}. The key idea is that information from expensive physics-based simulations has been distilled by machine learning and can be applied to new examples.}

@@

>>>>

A key new enabling technology breaks these classical tradeoffs in data-model comparisons for astronomical spectroscopy.  Automatic differentiation \citep[``autodiff'' or ``autograd'',][]{2015arXiv150205767G, 2016PhDT.......317M} and its affiliated backpropagation algorithm \citep{kelley1960,Linnainmaa1976TaylorEO, 1986Natur.323..533R, 1990JGCD...13..926D} has revolutionized machine learning and neural network architecture design, and is increasingly applied in astrophysical data analysis contexts, \emph{e.g.} kernel phase coronography with \texttt{morphine} \citep{2021ApJ...907...40P,2021JOSAB..38.2465W} \edit1{and $\partial$\texttt{lux} \citep{10.1117/12.2629774}}, \edit1{stellar surface modeling with \texttt{starry} \citep{2021AJ....162..123L}}, and exoplanet orbit fitting with \texttt{exoplanet} \citep{2021JOSS....6.3285F}.  Of the spectroscopy frameworks mentioned above, the \texttt{TensorFlow}-based \citep{tensorflow2015-whitepaper} \texttt{wobble} and the \texttt{JAX}-based \citep{jax2018github} \texttt{ExoJAX} employ autodiff technology.  \texttt{wobble} treats each pixel as a tunable control point, producing $\sim10^5$ parameters for a modern stellar spectrum.  The \texttt{ExoJAX} framework has only $\sim$dozens of tunable parameters that describe the fundamental physical properties controlling a brown dwarf atmosphere.  These two autodiff-aware frameworks span the extreme ends of non-parametric and parametric modeling for spectroscopy.

@@

We initialize the cloned model with a series of preprocessing steps. We divide the entire spectrum by a black body $\mathsf{B}(\bm{\lambda}_S)$ of the same effective temperature $T_{\mathrm{eff}}$ as the model template. The resulting signal usually still has smooth wiggles around the continuum. An optional continuum flattening step ensures that subsequent spectral line finding steps get applied uniformly. This high-pass filtering step should be set to capture the genuine spectral shape, without over-fitting broad line wings such as those in deep hydrogen and sodium lines. Any high-pass filter will work, a Gaussian Process approach would be ideal \citep{czekala15}.  Instead we apply a simple and familiar heuristic: fit a polynomial $\mathsf{P}(\bm{\lambda}_S)$ to a few continuum peaks and divide out the trend.

@@

\emph{``Surrogate modeling / Simulation Intelligence''}-- \edit1{Our slightly off-label use of machine learning straddles a boundary of machine learning and conventional physics-based models.  The rapidly evolving field of machine learning may therefore classify the approach here using a few related terms.  The phrases surrogate modeling, emulation, discrepancy modeling, transfer learning, simulation intelligence, and physics-informed machine learning could describe some aspects of \emph{blas\'e}, depending on the ultimate application by the end-user.  For example, machine learning has been applied to physical chemistry models using a neural network as a cheaper-but-still-accurate ``surrogate'' for the real theory-based simulations. The realized speedups in quantum chemistry can approach 300,000$\times$ \citep{pmlr-v70-gilmer17a}. The key idea is that information from expensive physics-based simulations has been distilled by machine learning and can be applied to new examples.  Scientific domain knowledge has been shown to aid in automatic exoplanet transit classification with Deep Learning \citep{2018ApJ...869L...7A}.}

@@

\subsection{PyTorch and the autodiff ecosystem}

\edit1{One key limitation stems from the need for autodiff to ``know about'' or sense each step of the end-to-end computation.  Practically speaking, this requirement means \texttt{blase} had to be written in a single differentiable (\emph{i.e.} autodiff-aware) programming framework.  This requirement is so strong as to forbid the intermingling of frameworks and languages.  The choice of which machine learning framework to adopt must therefore be made up-front, with some risk of \emph{vendor lock-in} once that choice has been set.  We considered three such differentiable frameworks: \texttt{TensorFlow}, \texttt{JAX}, and \texttt{PyTorch}.  We chose \texttt{PyTorch} because of its ease of use and its ability to structure differentiable models in an object-oriented style familiar within the \texttt{Python} community, with robust optimization support.  PyTorch had already been applied to several astronomy projects by this time, including \texttt{Exonet} \citep{2018ApJ...869L...7A}, \texttt{PySR} \citep{pysr}, and \texttt{MPoL} \citep{mpol}.  Ultimately, the choice of \texttt{PyTorch} made it easy to employ Sparse Matrices.}

\edit1{This vendor lock-in gives rise to a frustrating incompatibility among exoplanet-relevant autodiff applications.  For example, no differentiable piece of code in \texttt{blase} can call an \texttt{exoplanet} orbital radial velocity, a \texttt{starry} surface map, an \texttt{astropy} spectral coordinate transformation, a \texttt{wobble} spectral residual model, or an \texttt{ExoJAX} Voigt profile.  All of those tools use either \texttt{Theano}, \texttt{TensorFlow}, \texttt{Numpy}, or \texttt{JAX}, which are incompatible with \texttt{PyTorch}.  We use \texttt{SciPy}, but only for preprocessing, before the differentiable model has entered.  Some heuristics exist for interchanging code among the frameworks, but they pose some implementation complexity and/or some performance hit.}

\edit1{It may appear attractive to coordinate the relatively small autodiff-for-astrophysics community to adopt a single autodiff-able machine learning framework to facilitate compatibility.  In practice such coordination may be ill-fated, since the rapid pace of machine learning development has not yet settled on a single best way forward.  Each framework offers some strengths and demerits, and exploring those tradeoffs will continue to yield new experimental approaches.  For example, \texttt{JAX} may offer speedups attributable to Just In Time (JIT) compilation, at the expense of a more rigid functional programming style.  Access to scalable Gaussian Process implementations may be another deciding factor \citep{2022arXiv220908940A}.}
